---
title: "Assignment 1"
subtitle: "DATA311 - Machine Learning"
author: "Rin Meng, 51940633"
toc: true
format:
  html:
    embed-resources: true
    df-print: paged
editor: visual
---

## **Exploratory / Pre-processing**

### Exercise 1

b\. Regression

c\. Supervised

### Exercise 2

```{r}

abalone <- read.csv(("abalone.data"))

str(abalone)

```

### Exercise 3

```{r}

hist(abalone$X15, xlab = "Abalone age", ylab = "Abalone count", main = "Histogram of abalone count vs abalone age")
```

We see that this is an observation of a skewed normally distributed histogram. This should not break any assumptions of the linear model because in a linear model, we should have a data that follows:

1.  **Linearity:** The relationship between the independent and dependent variables should be linear.

2.  **Independence**: The residuals should be independent.

3.  **Homoscedasticity**: The residuals should have constant variance at every level of the independent variable.

4.  **Normality**: The residuals should be approximately normally distributed.

### Exercise 4

```{r}

set.seed(51940633)

# count half of the rows
sampleSize <- floor(0.5 * nrow(abalone))

# split rows into 2 sets
set <- sample(seq_len(nrow(abalone)), size = sampleSize)

abaloneTrain <- abalone[set, ]
abaloneTest <- abalone[-set, ]

abaloneTrain
abaloneTest

```

## **Simple Linear Regression**

### **Exercise 5**

```{r}
abaloneTrain.lm <- lm(X15 ~ X0.455,data=abaloneTrain)
summary(abaloneTrain.lm)
```

Since the p value for the length coefficient is incredibly small at ($< 2 \times 10^-16$), we can safely say that there is a significant relationship between abalone's length and rings.

### Exercise 6

The coefficient of determination, $R^2$ , also known as the Multiple R-squared value, is at $0.2846$. This therefore, means that about $28.46%$ of the presented abalone's ring is explainable by the abalone's length.

### Exercise 7

```{r}
plot(abaloneTrain$X0.455, abaloneTrain$X15, xlab="Abalone length (mm)", ylab="Abalone rings"
     ,pch=16,cex=0.3, main = "Scatter plot of rings vs length")
abline(abaloneTrain.lm, col = "red")

```

1.  **Linearity:** since the relationship between the independent variable (length), and the dependent variable (rings), appears to scatter in straight lines, making patterns, then linearity is generally considered as being satisfied.

2.  **Homoscedasticity:**

### Exercise 8

```{r}
qqnorm(residuals(abaloneTrain.lm), main="Q-Q Plot of Residuals", pch=16, cex=0.3)
qqline(residuals(abaloneTrain.lm), col="red")
hist(residuals(abaloneTrain.lm), breaks=30, main="Histogram of Residuals", xlab="Residuals", col="lightblue")
```

The normality of the residuals is not satisfied because the historgrams shows that it is heavily skewed, and the quantile-quantile plot shows that it does not fall within the quantile-quantile line.

## **Multiple Linear Regression**

### Exercise 10
